Nested sampling with two objective functions

### 2016-03-10 Paper/ms.tex
Working on the beginning of the paper

\DeclareUnicodeCharacter{00A0}{ }
\newcommand{\xx}{\boldsymbol{x}}
\newcommand{\dx}{d\boldsymbol{x}}
\newcommand{\data}{\boldsymbol{D}}
\newcommand{\II}{\boldsymbol{I}}

@@

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We present a generalization of the Nested Sampling algorithm for
estimating the partition function $Z(\beta_1, \beta_2)$ of a family of
probability distributions indexed by two hyperparameters $\beta_1$ and
$\beta_2$. These
problems arise frequently in statistical mechanics and occasionally in
Bayesian inference. We demonstrate the algorithm on three example problems.
The first is a simple toy example where
$Z(\beta_1, \beta_2)$ is known, where we show empirically how the accuracy
improves as a function of the number of particles $N$. We also
calculate the partition function for a collection of atoms with a potential
that is `intermediate' between a Lennard-Jones form and a polymer form, in
a single run. Finally, we present a Bayesian image reconstruction example,
where we can infer the value of a hyperparameter even though its prior
distribution contains an `intractable' normalizing constant.}

@@

Nested Sampling \citep[NS][]{skilling} is popular general
Monte Carlo algorithm that can solve a wide range of problems in Bayesian
computation and statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
Its key strength its ability to cope with phase transitions which cause
many other methods to fail.
In a Bayesian inference framework with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $\II$) is:
\begin{eqnarray}
p(\xx | \data, \II) &=&
\frac{p(\xx | \II)p(\data | \xx, \II)}{p(\data | \II)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, usually known as the
`marginal likelihood' or the `evidence':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.\label{eqn:evidence}
\end{eqnarray}

@@

NS works by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by a (roughly) known
and constant factor. Like related algorithms such as those
based on `annealing' or `tempering', NS
moves through a sequence of probability distributions beginning from the
prior. This sequence of distributions is defined by
\begin{eqnarray}
p(\xx; \ell) &=& \frac{1}{X(\ell)}\pi(\xx)\mathds{1}\left(L(\xx) > \ell\right).
\label{eq:constrained_prior}
\end{eqnarray}
and the marginal likelihood can be computed by numerically integrating the
function $L(X)$:
\begin{eqnarray}
Z &=& \int_0^1 L(X)\, dX.
\end{eqnarray}
Throughout this paper we use the popular `overloaded' notation for some
functions and probability distributions.
For example, $L(\xx)$ is the likelihood function, and $L(X)$ is the likelihood
as a function of the enclosed prior mass $X$. These cannot literally be the
same function as the inputs are not even necessarily of the same dimensionality.
Therefore the symbol $L$ implies that the output of the function is a
likelihood value; so $L(\xx)$ represents a likelihood value computed from
the parameters $\xx$, and $L(X)$ represents a likelihood value that
corresponds to an amount $X$ of prior mass.

Naively, Equation~\ref{eqn:evidence} appears straightforward to compute because
it is a simple expectation of $L$ with respect to the distribution $\pi$.
However,
NS is required because Equation~\ref{eqn:evidence}
is the expected value of $L$ with respect to a very heavy-tailed distribution
(the prior for the $L$ value implied by $\pi(\xx)$). Hence, there is a strong
connection between NS and ideas from rare event simulation, which has been
explicitly investigated by \citet{walter}.

>>>>

\DeclareUnicodeCharacter{00A0}{ }
\newcommand{\xx}{\boldsymbol{\theta}}
\newcommand{\dx}{d\boldsymbol{\theta}}
\newcommand{\data}{\boldsymbol{D}}
\newcommand{\II}{\boldsymbol{I}}

@@

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{We present a generalization of the Nested Sampling algorithm,
which we call ``{\em TwinPeaks}'',
that estimates
the partition function $Z(\beta_1, \beta_2)$ for a family of
probability distributions indexed by two hyperparameters, $\beta_1$ and
$\beta_2$. Problems like this
arise frequently in statistical mechanics and occasionally in
Bayesian inference.
We demonstrate the algorithm on three example problems.
The first is a toy problem whose structure allows us
to accurately calculate $Z(\beta_1, \beta_2)$ numerically using non-Monte-Carlo
methods.
We show empirically how the accuracy of TwinPeaks
improves as a function of the number of particles used, $N$. We also
calculate the partition function for a collection of atoms with a potential
that is `intermediate' between a Lennard-Jones form and a polymer form, in
a single run. Finally, we present a Bayesian image reconstruction example,
where we can infer the value of a hyperparameter even though its prior
distribution contains an `intractable' normalizing constant.}

@@

Nested Sampling \citep[NS][]{skilling} is popular general
Monte Carlo algorithm that can solve a wide range of problems in Bayesian
computation and statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
Its key strength its ability to cope with phase transitions which cause
problems for many other methods \citep{skilling}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $\II$) is:
\begin{eqnarray}
p(\xx | \data, \II) &=&
\frac{p(\xx | \II)p(\data | \xx, \II)}{p(\data | \II)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, usually known as the
`marginal likelihood' or the `evidence':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.\label{eqn:evidence}
\end{eqnarray}

@@

NS works by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by a (roughly) known
and constant factor. Like related algorithms such as those
based on `annealing' or `tempering', NS
moves through a sequence of probability distributions beginning from the
prior. This sequence of distributions is defined by
\begin{eqnarray}
p(\xx; \ell) &=& \frac{1}{X(\ell)}\pi(\xx)\mathds{1}\left(L(\xx) > \ell\right).
\label{eq:constrained_prior}
\end{eqnarray}
and the marginal likelihood can be computed by numerically integrating the
function $L(X)$:
\begin{eqnarray}
Z &=& \int_0^1 L(X)\, dX.
\end{eqnarray}
Throughout this paper we use the popular `overloaded' notation for some
functions and probability distributions.
For example, $L(\xx)$ is the likelihood function, and $L(X)$ is the likelihood
as a function of the enclosed prior mass $X$. These cannot literally be the
same function as the inputs typically don't have the same dimensionality
($\xx$ is usually a collection of quantities and $X$ is a single real number
in $[0, 1]$).
Therefore the symbol $L$ implies the output of the function is a
likelihood value; so $L(\xx)$ represents a likelihood value computed from
the parameters $\xx$, and $L(X)$ represents a likelihood value that
corresponds to an amount $X$ of prior mass.

Naively, Equation~\ref{eqn:evidence} appears straightforward to compute because
it is a simple expectation of $L$ with respect to the distribution $\pi$.
However,
NS is required because Equation~\ref{eqn:evidence}
is the expected value of $L$ with respect to a very heavy-tailed distribution
(the prior for the $L$ value implied by $\pi(\xx)$). Hence, there is a strong
connection between NS and ideas from rare event simulation. \citet{walter} has
explicitly explored this connection.
